---
title: "Liverpool Covid Model"
author: "Alex Phillips"
date: "10/3/2021"
output: 
  html_document
    
---

# Model overview


## Problem description

These models attempt to model the relationship between the numbers of phone calls to NHS 111 and/or NHS 111 online assessments reporting COVID-19 symptoms, and weekly recorded numbers of deaths from COVID-19.

## Model description

## Data 

An R script for generating some simulated data (based on the priors in the model) is in `src/scripts/generateSimulatedData.R` which can be used as follows:
```{r}
source("src/scripts/generateSimulatedData.R")
generatedData <- generateSimulatedData(
  seed = 0,
  maxTime = 100,
  population = 1000000,
  n_beta_pieces = 20,
  n_rho_calls_111_pieces = 10,
  calls_111_start = 30
)

stan_data <- generatedData$stan_data
ground_truth <- generatedData$ground_truth

```

This data can then be fit to the deaths and calls model:

```{r eval = F}
library(rstan)
options(mc.cores = parallel::detectCores()).
rstan_options(auto_write = TRUE)

initf <- function() {
  n_beta_pieces <- stan_data$n_beta_pieces
  
  initial_state_raw = c(runif(1, min = 0.99999, max = 1.0), runif(1, min = 0.0, max = 1.0))
  beta_left = exp(runif(n_beta_pieces, min = -2, max = 0.5))
  beta_right = exp(runif(n_beta_pieces, min = -2, max = 0.5))
  dL = runif(1, min = 3.5, max = 4.0)
  dI = runif(1, min = 2.2, max = 2.6)
  dT = runif(1, min = 11.0, max = 13.0)
  omega = 1/(1 + exp(-runif(1, min = -5, max = -3)))
  
  list(
    initial_state_raw = initial_state_raw,
    beta_left = beta_left,
    beta_right = beta_right,
    dL = dL,
    dI = dI,
    dT = dT,
    omega = omega
  )
}


# generate a list of lists to specify initial values
liverpoolModel <- stan_model("src/models/deaths_and_111_calls.stan")
fit <- sampling(liverpoolModel, data = stan_data, init = initf)
```

## Data generating process


### Data munging and examination

## Stan program

The Stan models are located at `Liverpool/src/models`:

```{r echo=FALSE, message=FALSE, comment=NA}
stan_file <- "src/models/deaths_and_111_calls.stan"
lines <- readLines(stan_file)
cat(paste(lines, "\n", sep=""), sep="")
```

## Running model


## Model validation


### Run posterior diagnostics

### Prior predictive check

#### Plotting simulations from priors

### Parameter recovery with simulated data

### Posterior predictive check

### Cross validation

### SBC validation


