---
title: "Model reproduction checklist"
author: "Breck Baldwin"
date: "March 21, 2021"
output: 
  html_document:
    includes:
       in_header: _html/ga.html
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	results = "hide",
	error = FALSE,
	comment = ''
)
```

# Model overview

This document exists in the repository at [https://github.com/codatmo/Liverpool](https://github.com/codatmo/Liverpool) as `model_reproduction_checklist.Rmd` and rendered in html as [https://github.com/codatmo/Liverpool/model_reproduction_checklist.html](https://github.com/codatmo/Liverpool/model_reproduction_checklist.html). 

The sister page [https://codatmo.github.io/Liverpool/index.html](https://codatmo.github.io/Liverpool/index.html) describes the model more fully. 

This model is part of the CoDatMo (Co)vid (Dat)a (Mo)deling site (https://codatmo.github.io/) which is intended to replicate and make available important COVID models written in Bayesian modeling languages like Stan or PyMC3. 

### Validation checklist items 

* Model released on github: https://github.com/codatmo/Liverpool
* Problem description
* Model description: Research goals, references, supplementary description as necessary.
* Data
  + Data generating process
  + Data munging and examination
* Stan program
* Running model
  + Small data set to validate model execution (not done)
  + Run on availble data
  + Examine model output
* Model 
* Model validation
  + Report posterior diagnostics
  + Prior predictive check parameters/predictions
  + Parameter recovery with simulated data
  + Posterior predictive check
  + Parameter interpretation (not done)
  + Cross validation (not done)
  + Simulation based calibration (SBC) (not done)
* Compute environment

## Problem description

Model goal is to estimate COVID hospitalizations based on reports from 111 calls to government services and/or website reports. 

## Model description
See [https://codatmo.github.io/Liverpool/index.html](https://codatmo.github.io/Liverpool/index.html) for detailed description. 

## Data 

There is a script to generate simulated data. 
```{r}
source("src/scripts/generateSimulatedData.R")
generatedData <- generateSimulatedData(
  seed = 0,
  maxTime = 100,
  population = 1000000,
  n_beta_pieces = 20,
  n_rho_calls_111_pieces = 10,
  calls_111_start = 30
)
stan_data <- generatedData$stan_data
stan_data$compute_likelihood <- 1 # controls whether likelihood is computed
ground_truth <- generatedData$ground_truth
```

The script does a fair amount of data munging, see below, but the data analogous to production data are two data sources:

```{r results='show'}
deaths <- stan_data$deaths
calls_111 <- stan_data$calls_111
cat(paste(sprintf("Weekly COVID death count=%d, Days of 111_calls (7 days per weekly count)=%d", 
                  length(deaths),
                  length(calls_111))))
```

```{r}
library(ggplot2)
library(tidyr)
library(gridExtra)

day <- 1:length(calls_111)
data.df <- data.frame(calls_111,day)
data.df$deaths <- ifelse(data.df$day %% 7 == 0, deaths[data.df$day %/% 7], NA)

ggplot(data.df) +
  aes(x=1:length(calls_111), y=calls_111) +
  geom_line() -> p1

ggplot(data.df) +
  aes(x=1:length(calls_111), y=deaths) +
  geom_point() -> p2

grid.arrange(p1, p2, ncol=2)
```

### Data munging

The data undergo considerable processing before being ready to fit with Stan. There are 22 data parameters of which two are the `deaths` and `calls_111` shown above. The data elaboration proceeds as follows:

```{r}
deaths <- stan_data$deaths
calls_111 <- stan_data$calls_111

# below are constants
T <- 100
maxTime <- T
initial_time <- 0
n_disease_states <- 8
n_beta_pieces <- 20
calls_111_start <- 30
n_rho_calls_111_pieces <- 10
population <- 1e+06

beta_pieceLength <- ceiling(maxTime / n_beta_pieces)
beta_left_t <- seq(0, maxTime - 1, by = beta_pieceLength)
beta_right_t <- c(beta_left_t[2:n_beta_pieces], maxTime + 1)

real_data_length = length(beta_left_t) + length(beta_right_t) + 1
real_data = c(beta_left_t, beta_right_t, population)
integer_data = c(
  maxTime,
  length(beta_left_t),
  length(beta_right_t),
  length(beta_left_t),
  n_disease_states
)
integer_data_length <- length(integer_data)

rho_calls_pieceLength <-
  ceiling((maxTime - calls_111_start) / n_rho_calls_111_pieces)
rho_calls_111_left_t <-
  seq(calls_111_start, maxTime - 1, by = rho_calls_pieceLength)
rho_calls_111_right_t <-
  c(rho_calls_111_left_t[2:n_rho_calls_111_pieces], maxTime + 1)

times <- 1:T
deaths_length <- length(deaths)
deaths_starts <- seq(from = 1, to = deaths_length * 7, by = 7)
deaths_stops <- seq(from = 7, to = deaths_length * 7, by = 7)
calls_111_length <- length(calls_111)

stan_data2 <- list(
  initial_time = initial_time,
  n_beta_pieces = n_beta_pieces,
  beta_left_t = beta_left_t,
  beta_right_t = beta_right_t,
  T = maxTime,
  times = times,
  n_disease_states = n_disease_states,
  population = population,
  deaths_length = length(deaths),
  deaths_starts = deaths_starts,
  deaths_stops = deaths_stops,
  deaths = deaths,
  real_data_length = length(beta_left_t) + length(beta_right_t) + 1,
  real_data = c(beta_left_t, beta_right_t, population),
  integer_data_length = 5,
  integer_data = c(
    maxTime,
    length(beta_left_t),
    length(beta_right_t),
    length(beta_left_t),
    n_disease_states
  ),
  n_rho_calls_111_pieces = n_rho_calls_111_pieces,
  rho_calls_111_left_t = rho_calls_111_left_t,
  rho_calls_111_right_t = rho_calls_111_right_t,
  calls_111_start = calls_111_start,
  calls_111 = calls_111,
  calls_111_length = calls_111_length,
  compute_likelihood = 1
)

# test that data has been properly munged.
for (varname in names(stan_data)) {
  # print(paste("trying:",varname))
  if (!varname %in% names(stan_data2)) {
    print(paste("NULL for ", varname))
  }
  if (!all(stan_data[[varname]] == stan_data2[[varname]])) {
    print(paste("mismatch ", varname))
  }
}
```

## Stan program

The Stan model is located at `src/models/deaths_and_111_calls.stan`:

```{r results='show'}
stan_file <- "src/models/deaths_and_111_calls.stan"
lines <- readLines(stan_file)
cat(paste(lines, "\n", sep=""), sep="")
```

## Running model

First some inits are setup:
```{r}
set.seed(0)
initf <- function() {
  n_beta_pieces <- stan_data$n_beta_pieces
  
  initial_state_raw = c(runif(1, min = 0.99999, max = 1.0), runif(1, min = 0.0, max = 1.0))
  beta_left = exp(runif(n_beta_pieces, min = -2, max = 0.5))
  beta_right = exp(runif(n_beta_pieces, min = -2, max = 0.5))
  dL = runif(1, min = 3.5, max = 4.0)
  dI = runif(1, min = 2.2, max = 2.6)
  dT = runif(1, min = 11.0, max = 13.0)
  omega = 1/(1 + exp(-runif(1, min = -5, max = -3)))
  
  list(
    initial_state_raw = initial_state_raw,
    beta_left = beta_left,
    beta_right = beta_right,
    dL = dL,
    dI = dI,
    dT = dT,
    omega = omega
  )
}

```

The code is run as:

```{r results='show'}
library(cmdstanr)
library(rstan)
library(xfun)

model <- cmdstan_model(file.path("src","models","deaths_and_111_calls.stan"))
num_warmup <- 1000
num_samples <- 1000
num_chains <- 4

fit_1 <- 
  xfun::cache_rds({model$sample(data=stan_data2, 
                                seed=999, 
                                chains=num_chains, 
                                init = initf, 
                                num_warmup=num_warmup, 
                                num_samples=num_samples,
                                output_dir="output")},
              rerun=FALSE, file="fit1", dir="_cache/")

rs_fit_1 <- rstan::read_stan_csv(fit_1$output_files())

print(fit_1$time())
```

#running some warmup and just 100 draws with two chains. 

Viewing a text based summary of the fit, this will be very long, it is commented out and not run:

```{r}
# fit_1$cmdstan_summary()
```

## Model validation

Below are the diagnostics used to help validate the model. 

### Run posterior diagnostics

There are standard diagnostics that look for errors in the posterior.

```{r}
fit_1$cmdstan_diagnose()
```

* Treedepth warnings 
* Divergence check passed
* E-BFMI satisfactory
* R-hat values satisfactory

### Prior predictive check

The prior predictive check estimates the model parameters without the likelihood being used. The resulting draws are then used to predict new data via predictive application of the likelihood given the draws. Note that `compute_likelihood = 0` prevents the likelihood being computed in the model fitting step. 

```{r}
stan_data_no_likelihood <- stan_data2
stan_data_no_likelihood$compute_likelihood <- 0

fit_2 <- xfun::cache_rds({model$sample(data=stan_data_no_likelihood, seed=999,
                                           chains=num_chains, init = initf, 
                                           num_warmup=num_warmup,
                                           num_samples=num_samples,
                                           output_dir="output")},
                             rerun=FALSE, file="fit2", dir="_cache/")

rs_fit_2 <- rstan::read_stan_csv(fit_2$output_files())
rs_ex_fit_2 <- rstan::extract(rs_fit_2)

random_draws <- sample(1:nrow(rs_ex_fit_2$pred_deaths), 10, replace=FALSE)
deaths_draws <- t(rs_ex_fit_2$pred_deaths[random_draws,])
deaths_draws.df <- data.frame(deaths_draws) 
names(deaths_draws.df) <- random_draws
draw_names <- colnames(deaths_draws.df)
weeks <- 1:nrow(deaths_draws.df)
p_data2.df <- cbind(deaths,weeks,deaths_draws.df)

p_long_data <- gather(p_data2.df, draw, deaths_sim, draw_names)

p3 <- ggplot(data=p_long_data, aes(x=weeks)) +
            geom_line(aes(y=deaths_sim, group=draw, color=draw), size=.5) +
            geom_line(aes(y=deaths), color="black", size=.5)
p3
```

Above we see the posterior of the priors without seeing any data. Generally some justification for prior distributions is expected. The above are weakly informative in that they cover a broad range of plausible values. 

```{r}
call_draws <- t(rs_ex_fit_2$pred_calls_111[random_draws,]) #keep same draws
call_draws.df <- data.frame(call_draws) 
names(call_draws.df) <- random_draws
days <- 1:nrow(call_draws.df)
p_data3.df <- cbind(calls_111,days,call_draws.df)

p_long_data3 <- gather(p_data3.df, draw, calls_sim, draw_names)

p4 <- ggplot(data=p_long_data3, aes(x=days)) +
            geom_point(aes(y=calls_sim, group=draw, color=draw), size=.5) +
            geom_line(aes(y=calls_111), color="black", size=.5)
p4
```

### Parameter recovery with simulated data

Parameter recovery establishes that for some small set of values the model reasons properly. We pick a draw from the above distributions, simulate data with it and then attempt to recover the parameters that we simulated with. We can look at the above graph and pick expected outliers or close to actual data. For this we pick 42 as a middle-of-the-road example. 

```{r}
library(bayesplot)
#Pick one arbitrary draw from the prior distribution
draw <- 42
stan_data_3 <- stan_data
stan_data_3$calls_111 <- rs_ex_fit_2$pred_calls_111[42,]
stan_data_3$deaths <- rs_ex_fit_2$pred_deaths[42,]
stan_data_3$compute_likelihood <- 1

fit_3 <- xfun::cache_rds({model$sample(data=stan_data_3, seed=999,
                                           chains=num_chains, init = initf, 
                                           num_warmup=num_warmup,
                                           num_samples=num_samples,
                                           output_dir="output")},
                             rerun=FALSE, file="fit3", dir="_cache/")

rs_fit_3 <- rstan::read_stan_csv(fit_3$output_files())
rs_fit_3_draws <- rstan::extract(rs_fit_3)

mcmc_intervals(fit_3$draws(variables=c("dL", "dI", "dT","omega")))
```

A subset of the estimated paramters shown, mean, 50% and 90% intervals indicated. Actual values below:

```{r results='show'}
report <- paste(sprintf("Draw number %d", draw),
sprintf("actual dL=%.2f", rs_ex_fit_2$dL[draw]), 
sprintf("actual dI=%.2f", rs_ex_fit_2$dI[draw]), 
sprintf("actual dT=%.2f", rs_ex_fit_2$dT[draw]), 
sprintf("actual omega=%.2f", rs_ex_fit_2$omega[draw]), 
sep="\n")
cat(report)
```

### Posterior predictive check

Compare predicted data from original run that used actual data.

#### Predict deaths
 
```{r}
rs_fit_1_draws <- rstan::extract(rs_fit_1)

summary.df <- as.data.frame(summary(rs_fit_1,pars=c("pred_deaths"), probs=c(.05,.5,.95))$summary)
weeks <- 1:length(deaths)
all.df <- cbind(summary.df, deaths, weeks) 
colnames(all.df) <- make.names(colnames(all.df)) # to remove % in the col names

p4 <- ggplot(all.df, mapping = aes(x = weeks)) +
  geom_ribbon(aes(ymin = X5., ymax = X95.), fill = "orange", alpha = 0.6) +
  geom_line(mapping = aes(y = X50.)) +
  geom_point(mapping = aes(y = deaths)) +
  labs(x = "Week", y = "Deaths")
p4
```

Actual deaths are black dots, mean predicted deaths black line, orange ribbon spans central 95% of draws.

#### Predict 111 calls

```{r }
rs_fit_1_draws <- rstan::extract(rs_fit_1)

summary_2.df <- as.data.frame(summary(rs_fit_1,pars=c("pred_calls_111"), probs=c(.05,.5,.95))$summary)
days <- 1:length(calls_111)
all_2.df <- cbind(summary_2.df, calls_111, days) 
colnames(all_2.df) <- make.names(colnames(all_2.df)) # to remove % in the col names

p5 <- ggplot(all_2.df, mapping = aes(x = days)) +
  geom_ribbon(aes(ymin = X5., ymax = X95.), fill = "green", alpha = 0.6) +
  geom_line(mapping = aes(y = X50.)) +
  geom_point(mapping = aes(y = calls_111)) +
  labs(x = "Day", y = "Calls 111")
p5
```

Actual calls are black dots, mean predicted calls black line, green ribbon spans central 90% of draws.

### Parameter interpretation

Impact of param changes on predictions.

### Cross validation

### SBC validation

### Computing environment
```{r results='show'}
devtools::session_info()
```

